### 2022-04-25 paper.tex
initial commit

\documentclass{IEEEtran}
\author{Henry Pick}
\title{Harmonic Analysis on the Symmetric Group Applied to Image Recognition}
\begin{document}
\maketitle
\begin{abstract}
    The branch of image recognition, 
\end{abstract}
\section{Motivation}
\section{}
\end{document}

### 2022-05-06 paper.tex
started on compression example

@@
-\documentclass{IEEEtran}
-\author{Henry Pick}
-\title{Harmonic Analysis on the Symmetric Group Applied to Image Recognition}
+\documentclass[12pt,technote]{IEEEtran}
+\usepackage{cite}
+\usepackage{amsmath}
+\usepackage[hidelinks]{hyperref}
+\usepackage{bookmark}
+\usepackage{amssymb}
+
+\author{Henry Pick, MATH189J}
+\title{Algebraic Signal Processing Theory in Image Processing}
+\date{3 April 2022}
 \begin{document}
@@
 \begin{abstract}
-    The branch of image recognition, 
+    In this paper, we explore constructs in image processing through the work of Markus P\"uschel on algebraic signal processing theory.
+His main objective was to introduce well-established concepts in signal processing in terms of a comprehensive algebraic theory.
+Here, we do much of the same, in fact a significant portion of this paper is devoted to reviewing essential ideas in his paper.
+We draw connections between abstract algebra and representation theory and the core concepts learned in many systems engineering classes.
+In my own effort to make this work original, I will explore several applied cases not covered in P\"uschel's work.
+As of now, my goal is to look at image compression techniques that use spectral decomposition
 \end{abstract}
-\section{Motivation}
-\section{}
+\section{Introduction}
+In nearly all fields of the applied sciences, we encounter a broad field called ``signal processing''.
+The term was traditionally used in electrical engineering but has now found use in a wide range of other fields including image processing, wireless communication, financial analysis, and machine learning.
+
+In the broadest sense, signal processing deals with functions defined on sets, often times countable sets in the context of digital computing.
+We then have a set of functions, commonly referred to as ``filters'' that transform these signals.
+In this paper, we focus on linear signal processing, which means that this set of functions operates on signals \textit{linearly}.
+That is, given a transformation $A$, signals $x$ and $y$, and scalar multiples $r_1$ and $r_2$ it holds that $A(r_1x + r_2y) = r_1A(x) + r_2A(y)$.
+
+To one with experience in one of the fields of signal processing, reading this last sentence should have been fairly straightforward, drawing on some assumptions.
+One reading this in the context of processing something like software defined radio (SDR) signals might have assumed that $x$ and $y$ live in the vector space $\mathbb{C}^n$ over the field $\mathbb{C}$ and that the expression $r_1x + r_2y$ is simply the vector sum of $x$ and $y$ scaled by $r_1$ and $r_2$.
+Reading this last sentence from an abstract algebra perspective should have seemed slightly awkward, however.
+Since we have not given an algebraic basis for these objects $A, x, y, r_1,$ and $r_2$, we have no conception of how these operations between objects are defined, how we might represent these objects linearly, and what it means for $A$ to operate on a signal.
+The instinct of the mathematician is to establish a comprehensive theory that encompasses all of signal processing and that is exactly what Markus P\"uschel has done in his work on algebraic signal processing theory\cite{AlgebraicSignalProcessing2006}.
+
+One might question the practical value of an algebraic exploration of an already well-established theory, in this we find many justifications.
+For example, an engineer who has significant experience in one field of signal processing will find a natural means of recognizing parallels in a totally different context.
+Furthermore, we find that certain algorithms that seem arbitrary in a standard perspective fall naturally out of this new theory.
+We will see this in a practical application of our learnings.%todo: reword this last paragraph
+\section{From First Principles}
+
+The most central concept in signal processing theory is the \textit{signal processing model}.
+This is defined as algebra $\mathcal{A}$ of filters, an $\mathcal{A}$-module $\mathcal{M}$ of signals, and a bijective mapping $\Psi$ from the vector space $V$ to the module $\mathcal{M}$ of signals.
+We often refer to this with an ordered triple $(\mathcal{A,M,\phi})$.
+We will see why this model is necessitated by first principles in the next section, but currently we
+
+\section{From First Principles}
+
+\subsection{Shifts and the Z-Transform}
+
+\cite{AlgebraicSignalProcessing2006}
+Basic constructs in signal processing theory are filters, the z-transform, signals and the fourier transform.
+When we view them as they are classically taught, we see filters as linear operators and signals as a vector space, but this paper employs an algebraic perspective to show that these are more than
+
+Z-transform converts a discrete time signal  into a complex frequency-domain representation, which can be thought of as the discrete time analogue of the Laplace transform. The
+
+Define this notion of ``linear time invariance''
+
+Introduce transfer functions in a formal context.
+We see that transfer functions are polynomials in $z^{-n}$, where $n$ is the order of the filter. For example, a third-order Butterworth filter is
+\begin{equation*}
+    H(s) = \frac{G_0}{B_n(a)} \qquad a = \frac{\omega_n}{s}
+\end{equation*}
+and $B_n$ is the Butterworth polynomial defined for even numbers as
+\begin{equation*}
+    B_n(s) = \prod_{k=1}^\frac{n}{2}
+\end{equation*}
+
+However, simple observation shows that this space is not closed under multiplication, the operation corresponding with successive filter application.
+As such, many engineering conventions treat this as multiplication of polynomials modulo $z^{-n}-1$
+
+\subsection{Signal Shift Operator}
+Every signal model that has the shift invariance property has a commutative filter algebra $\mathcal{A}$.
+\section{Transformations between Algebras}
+Many times in engineering we encounter examples where a transformation may not transform to the same algebra and may not be invertible.
+For example, the process of downsampling a signal (commonly referred to as applying a decimating filter) applies a transformation that reduces the number of samples in a vector.
+This can be thought of as applying a filter with a large null space to a signal.
+
+\section{Practical Example: JPEG Image Compression}
+\bibliographystyle{plain}
+\bibliography{paper}
 \end{document}

### 2022-05-06 src/compress.c

@@ -1,0 +1,110 @@
+/*
+Henry Pick 2022
+Harvey Mudd College
+hpick@hmc.edu
+
+DFT image compression using my own implementation of the discrete cosine transform
+
+This is intended to demonstrate the feasibility of using a ground-up approach 
+to develop robust signal processing applications with only a signal model and
+a basic engineering objective. I intentionally did not use the fftw3 library 
+for this reason, but in most contexts you would probably want to use a
+standard fft library for readability.
+*/
+
+#include <stdio.h>
+#include <string.h>
+#include <stdlib.h>
+#include <stdint.h>
+#include <fftw3.h>
+
+#define COMPRESSION_RATE 0.5
+
+struct pixel {
+    uint8_t red;
+    uint8_t green;
+    uint8_t blue;
+};
+
+typedef struct pixel * image;
+
+// read a ppm file and return an array of pixel structs
+int load_ppm(char *file_name, image *im, int *rows, int *cols)
+{
+    char type[3];
+    FILE *file = fopen(file_name, "rb");
+    int numRow, numCol, maxNum;
+
+    struct pixel currentPix;
+
+    char buf[5];
+
+    if (file == NULL) {
+        printf(stderr, "Could not open file %s for reading");
+        return -1;
+    }
+
+    fscanf(file, "%2s", type);
+    type[2] = '\0';
+    if(strcmp(type, "P6") != 0) {
+        printf(stderr, "This file is not of type P6");
+        return -1;
+    }
+
+    fscanf(file, "%d", &numCol);
+    fscanf(file, "%d", &numRow);
+    fscanf(file, "%d", &maxNum);
+
+    *rows = numRow;
+    *cols = numCol;
+
+    fseek(file, 1, SEEK_CUR);
+
+    //fread(&currentPix, sizeof(struct pixel), 1, file);
+
+    *im = (image)malloc(numRow * numCol * sizeof(struct pixel));
+    if(im == NULL) {
+        printf("memory not allocated!");
+    }
+
+    for (int i = 0; i < numCol * numRow; i++) {
+        fread(*im + i, sizeof(struct pixel), 1, file);
+    }
+
+    fclose(file);
+    
+    return 0;
+}
+
+#define BLOCK_SIZE 8
+
+// run dct on a given block and return the compression subset
+int dct_compress() {
+    // create float buffer and run dft2d on the blocks
[…]

### 2022-05-09 paper.tex
working image read and write functions

--- 2022-04-25 paper.tex
@@
-\documentclass{IEEEtran}
-\author{Henry Pick}
-\title{Harmonic Analysis on the Symmetric Group Applied to Image Recognition}
+\documentclass[12pt,technote]{IEEEtran}
+\usepackage{cite}
+\usepackage{amsmath}
+\usepackage[hidelinks]{hyperref}
+\usepackage{bookmark}
+\usepackage{amssymb}
+\usepackage{graphicx}
+
+\author{Henry Pick, MATH189J}
+\title{Algebraic Signal Processing Theory in Image Processing}
+\date{3 April 2022}
 \begin{document}
@@
 \begin{abstract}
-    The branch of image recognition, 
+    In this paper, we explore constructs in image processing through the work of Markus P\"uschel on algebraic signal processing theory.
+His main objective was to introduce well-established concepts in signal processing in terms of a comprehensive algebraic theory.
+Here, we do much of the same, in fact a significant portion of this paper is devoted to reviewing essential ideas in his paper.
+We draw connections between abstract algebra and representation theory and the core concepts learned in many systems engineering classes.
+In my own effort to make this work original, I will explore several applied cases not covered in P\"uschel's work.
+As of now, my goal is to look at image compression techniques that use spectral decomposition
 \end{abstract}
-\section{Motivation}
-\section{}
+\section{Introduction}
+In nearly all fields of the applied sciences, we interact with the broad definition of ``signal processing''.
+It is a term traditionally used in electrical engineering but has now found use in a wide range of other fields including image processing, wireless communication, financial analysis, and machine learning.
+
+In the broadest sense, signal processing deals with functions defined on sets, often times countable sets in the context of digital computing.
+We also have a set of functions, commonly referred to as ``filters'' that transform these signals.
+In this paper, we focus on linear signal processing, which means that this set of functions operates on signals \textit{linearly}.
+That is, given a transformation $A$, signals $x$ and $y$, and scalar multiples $r_1$ and $r_2$ it holds that $A(r_1x + r_2y) = r_1A(x) + r_2A(y)$, which is still a signal.
+
+To one with experience in one of the fields of signal processing, reading this last sentence should have been fairly straightforward, drawing on some assumptions.
+If one had contextualized this as the processing of complex digital sensor data, they might have assumed that $x$ and $y$ live in the vector space $\mathbb{C}^n$ over the field $\mathbb{C}$ and that the expression $r_1x + r_2y$ is simply the vector sum of $x$ and $y$ scaled by $r_1$ and $r_2$.
+Reading this last sentence from an abstract algebra perspective should have seemed slightly awkward, however.
+Since we have not given an algebraic foundation for these objects $A, x, y, r_1,$ and $r_2$, we have no conception of how the interact with one another, how we might represent these objects linearly, and what a signal generally represents.
+The instinct of the mathematician is to establish a comprehensive theory that characterizes features common to all signal processing contexts.
+This is exactly what Markus P\"uschel has done in his work on algebraic signal processing theory\cite{AlgebraicSignalProcessing2006}.
+
+One might question the practical value of an algebraic exploration of already well-established science.
+In this inquiry, we find many justifications that apply far beyond the mathematics alone.
+For example, an engineer who has significant experience in one field of signal processing may find a natural means of recognizing parallels in an entirely different field.
+It also becomes clear that several tasks in processing research that are seemingly the product of ingenious creativity are actually quite procedural.
+The design of fast transform algorithms, for example, can be as simple as observing the symmetry structure of an underlying signal module.
+
+The structure of this paper will first explore some of the foundational concepts introduced in the P\"uschel paper, which is primarily an expository section.
+In this section, we also consider some ideas that were not addressed in this work and potential extensions for P\"uschel's signal processing theory.
+We then move on to a practical application of constructing a signal model for images and a fast transform to be used in an image compression program.
+\section{From First Principles}
+In standard signal processing theory, signals live in a vector space like $\mathbb{C}^n$ over the base field, which gives the direct operations of signal addition and ``scalar multiplication''.
+This vector space structure also endows us with several other notions like dimension, basis, linear mappings and subspaces.
+We then encounter objects called filters, linear operators that act on signals to produce new signals.
+The operation of a filter on a signal is a different operation than the sum of two signals.
+We use the multiplication operation to represent this action of a filter on a signal
+\begin{equation*}
+    \text{filter}\cdot\text{signal} = \text{signal}
+\end{equation*}
+filtering also obeys the laws of distributivity
+\begin{align*}
+    &\text{filter}\cdot (\text{signal} + \text{signal}) \\
+    =\ &\text{filter}\cdot \text{signal} + \text{filter}\cdot\text{signal}
+\end{align*}
+For $h,h'\in \mathcal{A}$ and $s,s'\in\mathcal{M}$ and $\alpha$ in the base field, we also have 
+\begin{align*}
+    h\cdot (s + s') &= h\cdot s + h\cdot s'\\
+    h\cdot (\alpha s) &= \alpha (h\cdot s)\\
+    h'\cdot (h\cdot s) &= (h'\cdot h)\cdot s
+\end{align*}
+and filters can be operated on in various manners as well
+\begin{align*}
+    \text{filter} + \text{filter} &= \text{filter (parallel connection)}\\
+    \text{filter} \cdot \text{filter} &= \text{filter (series connection)}\\
+    \alpha\cdot\text{filter} &= \text{filter (amplification)}
+\end{align*}
+Beyond these properties, it must be mentioned that scalar multiplication of filters has left and right-distributivity and that multiplication by scalars is fully commutative with filter multiplication.
+At this point, we see that the structure of this system can be defined using formal mathematical language.
+Our set of filters forms an algebra, which we refer to as $\mathcal{A}$.
+Our vector space is somehow operated on by $\mathcal{A}$, meaning it must have an associated $\mathcal{A}$-module called $\mathcal{M}$.
+However, it must be clearly stated that $\mathcal{M}$ does not necessarily equal $V$ as the operation of $\mathcal{A}$ on $\mathcal{M}$ is not strictly defined as an operation on a vector space.
+Thus, we must also define a bijection between $V$ and $\mathcal{M}$, which we call $\Phi$.
+
+This is the central device in signal processing theory: a \textit{signal processing model} on a vector space $V$ consists of an ordered tuple $(\mathcal{A}, \mathcal{M}, \Phi)$ of algebraic objects as defined above.
+Through this model, we can exploit all of the mathematics of representation theory, harmonic analysis, and
+
+One may note that as soon as $\Phi$ is defined, we have also given definitions for all of $V, \mathcal{M}$ and $\mathcal{A}$, meaning a signal model can be tersely expressed as this bijective function $\Phi$.
+
+\subsection{Fourier Transforms}
+Given that we have now associated the signal space with a module, we can define Fourier transforms on our signal space.
+For an arbitrary $\mathcal{A}$-module $\mathcal{M}$, let $M$ denote the set of all irreducible $\mathcal{A}$-submodules.
+In the general case, we cannot be certain that $\mathcal{M}$ can be decomposed into $M$ but in the case that
+\begin{equation*}
+    \mathcal{M} = \bigoplus_{\omega\in M}\omega
+\end{equation*}
+we have the machinery to construct a Fourier Transform for our signal model.
+Once we choose bases for the irreducible representations of $\mathcal{M}$ allows us to coordinatize this transformation, thus offering a linear representation which we will call $\mathcal{F}$.
+
+\subsection{Shifts and the Z-Transform}
+
+Anyone who has focused on control systems or electronics in engineering is likely familiar with the concept of transfer functions on signals and filters represented in the z-domain.
+For example, the z-transform as it pertains to signals over infinite continuous spaces is often the bilateral Laplace transform
+\begin{equation*}
+    X(s) = \mathcal{L}\{x(t)\} = \int_{-\infty}^\infty x(t)e^{-st}
+\end{equation*}
+as we note, one of the convenient properties of such a transform is that the multiplication of signals and filters in the z-domain is simple polynomial multiplication.
+This property is often given without any other context, but here we can trivially justify why this is the case.
+
+\section{Space and Time Signals}
+
+\section{Deriving Transforms from a Signal Model}
+
+\cite{AlgebraicSignalProcessing2006}
+Basic constructs in signal processing theory are filters, the z-transform, signals and the fourier transform.
+When we view them as they are classically taught, we see filters as linear operators and signals as a vector space, but this paper employs an algebraic perspective to show that these are more than
+
+Z-transform converts a discrete time signal  into a complex frequency-domain representation, which can be thought of as the discrete time analogue of the Laplace transform. The
+
+Define this notion of ``linear time invariance''
+
+However, simple observation shows that this space is not closed under multiplication, the operation corresponding with successive filter application.
+As such, many engineering conventions treat this as multiplication of polynomials modulo $z^{-n}-1$
+
+\subsection{Signal Shift Operator}
+Every signal model that has the shift invariance property has a commutative filter algebra $\mathcal{A}$.
+\section{Further Considerations: Transformations between Algebras}
+Many times in engineering we encounter examples where a transformation may not transform to the same algebra and may not be invertible.
+For example, the process of downsampling a signal (commonly referred to as applying a decimating filter) applies a transformation that reduces the number of samples in a vector.
+This can be thought of as applying a filter with a large null space to a signal.
+
+\section{Practical Example: Spectral Image Compression}
+Data compression is a large subfield of information theory that involves the process of encoding information to use fewer bits than an original representation.
+Data compression usually comes at the expense of additional computation, which is why the development of efficient algorithms to compress data is a central focus of this field.
+Compressing a signal into its irreducible representations is one technique that is commonly used to compress digital data of many formats such as audio or images.
+In this section we look at defining a signal model for digital image data and then develop efficient transforms to represent images in a
+
+\subsection{Choosing an Encoding}
+There are a plethora of encodings for digital image data.
+Our goal here is to use an encoding that is uncompressed: one whose representation is as closest to the direct signal that one might encounter when processing raw image data from a sensor.
+We also want an encoding whose ``samples'' are easily accessible to a reading program.
+%TODO: better wording for this last sentence
+One family of encodings that stands out is the Netbpbm format, which encodes an image in a table of bytes $(0-255)$ in ASCII. It supports both grayscale and three-channel color representations.
+
+\subsection{Signal Model}
+
+\subsection{}
+\subsection{Demonstration}
+We first take an image and convert it into its uncompressed form, in this example, I have taken the following image and converted it into binary ppm:
+\begin{center}
+    \includegraphics[width=2in]{figures/lion.jpg}
+\end{center}
+\bibliographystyle{plain}
+\bibliography{paper}
 \end{document}

### 2022-05-09 src/compress.c

@@ -1,110 +1,0 @@
-/*
-Henry Pick 2022
-Harvey Mudd College
-hpick@hmc.edu
-
-DFT image compression using my own implementation of the discrete cosine transform
-
-This is intended to demonstrate the feasibility of using a ground-up approach 
-to develop robust signal processing applications with only a signal model and
-a basic engineering objective. I intentionally did not use the fftw3 library 
-for this reason, but in most contexts you would probably want to use a
-standard fft library for readability.
-*/
-
-#include <stdio.h>
-#include <string.h>
-#include <stdlib.h>
-#include <stdint.h>
-#include <fftw3.h>
-
-#define COMPRESSION_RATE 0.5
-
-struct pixel {
-    uint8_t red;
-    uint8_t green;
-    uint8_t blue;
-};
-
-typedef struct pixel * image;
-
-// read a ppm file and return an array of pixel structs
-int load_ppm(char *file_name, image *im, int *rows, int *cols)
-{
-    char type[3];
-    FILE *file = fopen(file_name, "rb");
-    int numRow, numCol, maxNum;
-
-    struct pixel currentPix;
-
-    char buf[5];
-
-    if (file == NULL) {
-        printf(stderr, "Could not open file %s for reading");
-        return -1;
-    }
-
-    fscanf(file, "%2s", type);
-    type[2] = '\0';
-    if(strcmp(type, "P6") != 0) {
-        printf(stderr, "This file is not of type P6");
-        return -1;
-    }
-
-    fscanf(file, "%d", &numCol);
-    fscanf(file, "%d", &numRow);
-    fscanf(file, "%d", &maxNum);
-
-    *rows = numRow;
-    *cols = numCol;
-
-    fseek(file, 1, SEEK_CUR);
-
-    //fread(&currentPix, sizeof(struct pixel), 1, file);
-
-    *im = (image)malloc(numRow * numCol * sizeof(struct pixel));
-    if(im == NULL) {
-        printf("memory not allocated!");
-    }
-
-    for (int i = 0; i < numCol * numRow; i++) {
-        fread(*im + i, sizeof(struct pixel), 1, file);
-    }
-
-    fclose(file);
-    
-    return 0;
-}
-
-#define BLOCK_SIZE 8
-
-// run dct on a given block and return the compression subset
-int dct_compress() {
-    // create float buffer and run dft2d on the blocks
[…]

### 2022-05-09 src/jhpeg.c

@@ -1,0 +1,182 @@
+
+#include "jhpeg.h"
+
+#include <stdio.h>
+#include <string.h>
+#include <math.h>
+#include <complex.h>
+#include <stdlib.h>
+#include <fftw3.h>
+
+#define BLOCK_SIZE 8
+
+
+// read a ppm file and return an array of pixel structs
+int load_ppm_rgb(char *file_name, image_rgb *im, int *rows, int *cols)
+{
+    char type[3];
+    FILE *file = fopen(file_name, "rb");
+    int numRow, numCol, maxNum;
+
+    char buf[5];
+
+    if (file == NULL) {
+        fprintf(stderr, "Could not open file %s for reading", file_name);
+        return -1;
+    }
+
+    fscanf(file, "%2s", type);
+    type[2] = '\0';
+    if(strcmp(type, "P6") != 0) {
+        printf(stderr, "This file is not of type P6");
+        return -1;
+    }
+
+    fscanf(file, "%d", &numCol);
+    fscanf(file, "%d", &numRow);
+    fscanf(file, "%d", &maxNum);
+
+    *rows = numRow;
+    *cols = numCol;
+
+    fseek(file, 1, SEEK_CUR);
+
+    *im = (image_rgb)malloc(numRow * numCol * sizeof(struct pixel_rgb));
+    if(im == NULL) {
+        printf("memory not allocated!");
+    }
+
+    for (int i = 0; i < numCol * numRow; i++) {
+        fread(*im + i, sizeof(struct pixel_rgb), 1, file);
+    }
+
+    fclose(file);
+    
+    return 0;
+}
+
+int write_ppm_rgb(char *file_name, image_rgb *im, int rows, int cols) {
+    FILE *file = fopen(file_name, "wb");
+
+    if (file == NULL) {
+        printf(stderr, "Could not open file %s for writing", file_name);
+        return -1;
+    }
+
+    //write file header
+    fprintf(file, "%2s\n%d %d\n%d\n", "P6", rows, cols, 255);
+
+    fwrite(im, sizeof(struct pixel_rgb), rows * cols, file);
+    fclose(file);
+    return 0;
+}
+
+
+// run dct on a given block and return the compression subset
+int dct_compress() {
+    // create float buffer and run dft2d on the blocks
+
+    return 0;
+}
+
+/*
+* for array size N, stride size s, and offset o, compute out-of-place 
+* cooley tukey in -> out
+*/
[…]

### 2022-05-09 src/jhpeg.h

@@ -1,0 +1,68 @@
+/*
+Henry Pick 2022
+Harvey Mudd College
+hpick@hmc.edu
+
+DFT image compression using my own implementation of the discrete cosine transform
+
+This is intended to demonstrate the feasibility of using a ground-up approach 
+to develop robust signal processing applications with only a signal model and
+a basic engineering objective. I intentionally did not use the fftw3 library 
+for this reason, but in most contexts you would probably want to use a
+standard fft library for readability.
+
+The output files are in a custom encoding, the "jhpeg" format, which is only meant
+to be read and decompressed by this program
+*/
+
+#include <stdint.h>
+#include <complex.h>
+
+// rgb pixel
+struct pixel_rgb {
+    uint8_t red;
+    uint8_t green;
+    uint8_t blue;
+};
+
+// bw pixel
+typedef uint8_t pixel_bw;
+
+struct compressed_rgb {
+    float red;
+    float green;
+    float blue;
+};
+
+
+typedef struct pixel_rgb* image_rgb;
+typedef struct pixel_bw* image_bw;
+
+
+typedef struct compressed_pix_rgb* compressed_image_rgb;
+typedef struct compressed_pix_bw* compressed_image_bw;
+
+/*
+* read ppm file and return pixel array
+*/
+int load_ppm_bw(char *file_name, image_bw *im, int *rows, int *cols);
+int load_ppm_rgb(char *file_name, image_rgb *im, int *rows, int *cols);
+
+
+/*
+* write ppm file from pixel array
+*/ 
+int write_ppm_rgb(char *file_name, image_rgb *im, int rows, int cols);
+
+/*
+* read in ppm at input_path and output compressed jhpeg format at output_path with float compression specifying the proportion of frequencies to remove
+*/
+int compress_ppm_bw(char *input_path, char *output_path, float compression);
+int compress_ppm_rgb(char *input_path, char *output_path, float compression);
+
+/*
+* read in compressed jhpeg at input_path, write ppm at output_path
+*/
+int decompress_jhpeg_bw(char *input_path, char *output_path);
+int decompress_jhpeg_rgb(char *input_path, char *output_path);
+

### 2022-05-09 src/jhpeg.c
not segfaulting

@@ -1,6 +1,3 @@
-
-#include "jhpeg.h"
-
 #include <stdio.h>
 #include <string.h>
 #include <math.h>
@@ -8,8 +5,25 @@
 #include <stdlib.h>
 #include <fftw3.h>
 
+#include "jhpeg.h"
+
 #define BLOCK_SIZE 8
 
+
+void create_compressed_im(struct compressed_im *compressed, int rows, int cols, uint8_t compression) {
+    int comp_size = 8 - compression;
+    int n = (ceil(rows / 8.0) * ceil(cols / 8.0))*comp_size*comp_size;
+
+    compressed->red = (int16_t*)malloc(n * sizeof(int16_t));
+    compressed->green = (int16_t*)malloc(n * sizeof(int16_t));
+    compressed->blue = (int16_t*)malloc(n * sizeof(int16_t));
+}
+
+void delete_compressed_im(struct compressed_im *compressed) {
+    free(compressed->red);
+    free(compressed->green);
+    free(compressed->blue);
+}
 
 // read a ppm file and return an array of pixel structs
 int load_ppm_rgb(char *file_name, image_rgb *im, int *rows, int *cols)
@@ -46,12 +60,10 @@ int load_ppm_rgb(char *file_name, image_rgb *im, int *rows, int *cols)
         printf("memory not allocated!");
     }
 
-    for (int i = 0; i < numCol * numRow; i++) {
-        fread(*im + i, sizeof(struct pixel_rgb), 1, file);
-    }
+    fread(*im, sizeof(struct pixel_rgb), numRow*numCol, file);
 
     fclose(file);
-    
+
     return 0;
 }
 
@@ -66,66 +78,195 @@ int write_ppm_rgb(char *file_name, image_rgb *im, int rows, int cols) {
     //write file header
     fprintf(file, "%2s\n%d %d\n%d\n", "P6", rows, cols, 255);
 
-    fwrite(im, sizeof(struct pixel_rgb), rows * cols, file);
+    fwrite(*im, sizeof(struct pixel_rgb), rows * cols, file);
+    fclose(file);
+    return 0;
+}
+
+int load_jhpeg_rgb(char *file_name, struct compressed_im *im, int *rows, int *cols, uint8_t *compression) {
+    FILE *file = fopen(file_name, "rb");
+    if (file == NULL) {
+        fprintf(stderr, "Could not open file %s for reading", file_name);
+        return -1;
+    }
+
+    // don't have any filetype safety, seems overkill
+    fscanf(file, "%d", rows);
[…]

### 2022-05-09 src/jhpeg.h

@@ -25,29 +25,26 @@ struct pixel_rgb {
     uint8_t blue;
 };
 
-// bw pixel
-typedef uint8_t pixel_bw;
+// struct channel_im {
+//     uint8_t *red;
+//     uint8_t *green;
+//     uint8_t *blue;
+// };
 
-struct compressed_rgb {
-    float red;
-    float green;
-    float blue;
+
+struct compressed_im {
+    int16_t *red;
+    int16_t *green;
+    int16_t *blue;
 };
 
-
 typedef struct pixel_rgb* image_rgb;
-typedef struct pixel_bw* image_bw;
 
-
-typedef struct compressed_pix_rgb* compressed_image_rgb;
-typedef struct compressed_pix_bw* compressed_image_bw;
 
 /*
 * read ppm file and return pixel array
 */
-int load_ppm_bw(char *file_name, image_bw *im, int *rows, int *cols);
 int load_ppm_rgb(char *file_name, image_rgb *im, int *rows, int *cols);
-
 
 /*
 * write ppm file from pixel array
@@ -55,14 +52,17 @@ int load_ppm_rgb(char *file_name, image_rgb *im, int *rows, int *cols);
 int write_ppm_rgb(char *file_name, image_rgb *im, int rows, int cols);
 
 /*
-* read in ppm at input_path and output compressed jhpeg format at output_path with float compression specifying the proportion of frequencies to remove
+* read ppm file and return pixel array
 */
-int compress_ppm_bw(char *input_path, char *output_path, float compression);
-int compress_ppm_rgb(char *input_path, char *output_path, float compression);
+int load_jhpeg_rgb(char *file_name, struct compressed_im *im, int *rows, int *cols, uint8_t *compression);
+
+/*
+* read in ppm at input_path and output compressed jhpeg format at output_path
+* with float compression specifying the proportion of frequencies to remove
+*/
+void compress_ppm(char *input_path, char *output_path, uint8_t compression);
 
 /*
 * read in compressed jhpeg at input_path, write ppm at output_path
 */
-int decompress_jhpeg_bw(char *input_path, char *output_path);
-int decompress_jhpeg_rgb(char *input_path, char *output_path);
-
+void decompress_jhpeg(char *input_path, char *output_path);

### 2022-05-09 src/jhpeg.c
compression/decompression/cli

@@ -1,3 +1,8 @@
+/* Working and not working
+* image reading and writing
+* fft forward and backward transforms on 2d arrays of small datatypes seems to work
+* compression and decompression works on square images, issues with different geometries
+*/
 #include <stdio.h>
 #include <string.h>
 #include <math.h>
@@ -45,6 +50,7 @@ int load_ppm_rgb(char *file_name, image_rgb *im, int *rows, int *cols)
         printf(stderr, "This file is not of type P6");
         return -1;
     }
+
 
     fscanf(file, "%d", &numCol);
     fscanf(file, "%d", &numRow);
@@ -196,9 +202,10 @@ void dct_compress(image_rgb *im, struct compressed_im *compressed, int rows, int
                         blue_in[j + i*8] = 255;
 
                     } else {
-                        red_in[j + i*8] = (*im)[c + j + (r+i)*cols].red;
-                        green_in[j + i*8] = (*im)[c + j + (r+i)*cols].green;
-                        blue_in[j + i*8] = (*im)[c + j + (r+i)*cols].blue;
+                        int pixel_index = j + c*8 + cols*8*r + cols*i;
+                        red_in[j + i*8] = (*im)[pixel_index].red;
+                        green_in[j + i*8] = (*im)[pixel_index].green;
+                        blue_in[j + i*8] = (*im)[pixel_index].blue;
                     }
                 }
             }
@@ -235,8 +242,13 @@ void dct_decompress(struct compressed_im *compressed, image_rgb *im, int rows, i
 
     for(int r = 0; r < comp_r; r++) {
         for(int c = 0; c < comp_c; c++) {
+            memset(red_in, 0, 64*sizeof(int16_t));
+            memset(blue_in, 0, 64*sizeof(int16_t));
+            memset(green_in, 0, 64*sizeof(int16_t));
+
             for(int i = 0; i < cell_size; i++) {
                 for (int j = 0; j < cell_size; j++) {
+
                     int compression_index =  (r*comp_c*cell_size*cell_size + i*comp_c*cell_size) + (c*cell_size + j);
 
                     red_in[i*8 + j] = compressed->red[compression_index];
@@ -256,9 +268,11 @@ void dct_decompress(struct compressed_im *compressed, image_rgb *im, int rows, i
                         continue;
 
                     } else {
-                        (*im)[c + j + (r+i)*cols].red = red_out[j + i*8];
[…]

### 2022-05-09 paper.tex
demo documentation

@@
 \usepackage{amsmath}
-\usepackage[hidelinks]{hyperref}
+\usepackage{hyperref}
 \usepackage{bookmark}
@@
 \usepackage{graphicx}
+\usepackage{caption}
+\usepackage{subcaption}
+\usepackage{hyperref}
+
+\hypersetup{
+    colorlinks=true,
+    linkcolor=blue,
+    filecolor=magenta,      
+    urlcolor=cyan,
+    pdfpagemode=FullScreen,
+}
 
@@
 We draw connections between abstract algebra and representation theory and the core concepts learned in many systems engineering classes.
-In my own effort to make this work original, I will explore several applied cases not covered in P\"uschel's work.
-As of now, my goal is to look at image compression techniques that use spectral decomposition
+In my own effort to make this work original, I will explore several applied cases not covered in P\"uschel's work and design an image compression pipeline that makes use of several of P\"uschel's discussion points.
 \end{abstract}
@@
 \section{Practical Example: Spectral Image Compression}
+The accompanying source code for this portion can be found \href{https://github.com/henry-2025/math189-spectral-compression}{here}.
 Data compression is a large subfield of information theory that involves the process of encoding information to use fewer bits than an original representation.
@@
 
-\subsection{}
 \subsection{Demonstration}
 We first take an image and convert it into its uncompressed form, in this example, I have taken the following image and converted it into binary ppm:
-\begin{center}
-    \includegraphics[width=2in]{figures/lion.jpg}
-\end{center}
+\begin{figure}
+    \centering
+    \begin{subfigure}[m]{0.3\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/lion_compressed.jpg}
+        \caption{A high-resolution image with 75\% compression applied}
+        \label{fig:lion}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.3\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/b_compressed.jpg}
+        \caption{A lower-resolution image with ``sharp'' features. Artifacts in the 75\% compression are more evident.}
+        \label{fig:letter_b}
+    \end{subfigure}
+    \label{fig:compression_outputs}
+    \caption{Two different outputs of the compression algorithm that demonstrate its performance.}
+\end{figure}
 \bibliographystyle{plain}

### 2022-05-09 src/jhpeg.c

@@ -47,7 +47,7 @@ int load_ppm_rgb(char *file_name, image_rgb *im, int *rows, int *cols)
     fscanf(file, "%2s", type);
     type[2] = '\0';
     if(strcmp(type, "P6") != 0) {
-        printf(stderr, "This file is not of type P6");
+        fprintf(stderr, "This file is not of type P6");
         return -1;
     }
 
@@ -77,7 +77,7 @@ int write_ppm_rgb(char *file_name, image_rgb *im, int rows, int cols) {
     FILE *file = fopen(file_name, "wb");
 
     if (file == NULL) {
-        printf(stderr, "Could not open file %s for writing", file_name);
+        fprintf(stderr, "Could not open file %s for writing", file_name);
         return -1;
     }
 
@@ -283,7 +283,7 @@ void dct_decompress(struct compressed_im *compressed, image_rgb *im, int rows, i
 
 
 void usage() {
-        printf("Usage:\tjhpeg -c INPUT OUTPUT COMPRESSION\tInput ppm binary encoding and output in jhpeg compression format\n\tjhpeg -d INPUT OUTPUT\t\t\tInput jhpeg compression file, output decompressed ppm\n");
+        printf("Usage:\tjhpeg -c INPUT OUTPUT COMPRESSION\tInput ppm binary encoding and output in jhpeg compression format\n\tjhpeg -d INPUT OUTPUT\t\t\tInput jhpeg compression file, output decompressed ppm\n\n\tCOMPRESSION is an integer 0-7 that specifies the level of rank reduction we apply to the DCT outputs. 0 is the lowest (this will actually augment the file size) while 7 is the highest, resulting in 1/32 the original size.\n");
 }
 
 void delete_image(image_rgb *im) {

### 2022-05-10 paper.tex
added a few pages

@@
 \usepackage{hyperref}
+\usepackage{tikz}
+
+\usetikzlibrary{shapes.geometric, arrows}
 
@@
 \section{Practical Example: Spectral Image Compression}
-The accompanying source code for this portion can be found \href{https://github.com/henry-2025/math189-spectral-compression}{here}.
 Data compression is a large subfield of information theory that involves the process of encoding information to use fewer bits than an original representation.
-Data compression usually comes at the expense of additional computation, which is why the development of efficient algorithms to compress data is a central focus of this field.
-Compressing a signal into its irreducible representations is one technique that is commonly used to compress digital data of many formats such as audio or images.
-In this section we look at defining a signal model for digital image data and then develop efficient transforms to represent images in a
+Compressing data usually comes at the expense of additional computation, which is why the development of efficient compression algorithms is a central focus of this field.
+We focus specifically on lossy compression, in which information is lost through the compression processes (ie.
+the original data cannot be reconstructed from the compressed data) but the majority of useful information is preserved.
+One common lossy compression technique involves decomposing some signal into its irreducible representations and then discarding the less-meaningful representations.
+In this section we look at defining a signal model for digital image data and then develop a lossy compression technique based on an efficient transform derived from this signal model.
+There is also a source code accompaniment for this section which can be found \href{https://github.com/henry-2025/math189-spectral-compression}{here}.
 
@@
 There are a plethora of encodings for digital image data.
-Our goal here is to use an encoding that is uncompressed: one whose representation is as closest to the direct signal that one might encounter when processing raw image data from a sensor.
-We also want an encoding whose ``samples'' are easily accessible to a reading program.
-%TODO: better wording for this last sentence
-One family of encodings that stands out is the Netbpbm format, which encodes an image in a table of bytes $(0-255)$ in ASCII. It supports both grayscale and three-channel color representations.
+Our goal here is to convert between two formats, one compressed and the other uncompressed.
+The uncompressed format is closest to the representation of raw sensor signal data while the compressed format is like one that we might wish to store on the hard drive of a computer.
+One family of encodings that stands out is the Netbpbm format, which encodes an image in a byte table corresponding directly to the pixel indices in the displayed image.
+For example, using the using red green and blue channel ordering for three-tuples, the following pixel array and color block are equivalent representations
 
+\begin{center}
+    $\begin{bmatrix}
+        (255, 0, 0) & (0, 255, 0) & (0, 0, 255)\\
+        (255, 255, 0) & (255, 255, 255) & (0, 0, 0)
+    \end{bmatrix}$\\
+    $\downarrow$\\
+    \includegraphics[width=0.5in]{figures/ppm_example.png}
+\end{center}
+As such, we will refer to our signal as a set of two-dimensional arrays of 8-bit unsigned integers for each color channel, which establishes our vector space $V$ for signals.
+We will say that elements in $V$ look like $\mathbf{s} = (s_{k,l})$, 2D arrays of elements in $s_{k,l}\in \mathbb{Z}/ 256\mathbb{Z}$.
+It may not make practical sense to think of this as a vector space right now, given that addition of signals and scalar multiplication seem like unnecessary operations.
+However, we hold this definition for the following section in which we derive the surrounding signal model.
 \subsection{Signal Model}
+In the discussion of shift operators, we found that spatial shifts are a necessary characteristic of data that does not have implicit directionality.
+We observed that for one dimensional signals, there exists only one shift operator that propagates outward from a single point in space.
+Here, we might intuitively think that one shift operator is also sufficient by just assigning the operator
+\begin{equation*}
+    q \diamond s_{x,y} = \frac{1}{4}(s_{x-1,y} + s_{x+1,y} + s_{x,y-1} + s_{x,y+1})
+\end{equation*}
+However, we soon realize see that this model would be flawed because we cannot reach all points in a 2d array from a single spatial operator.
+Instead, we define two spatial shifts, $q_x$ and $q_y$ and assign their propagations to each of the two dimensions such that
+\begin{align*}
+    q_x \diamond s_{x,y} &= \frac{1}{2}(s_{x-1,y} + s_{x+1,y})\\
+    q_y \diamond s_{x,y} &= \frac{1}{2}(s_{x,y-1} + s_{x,y+1})
+\end{align*}
+Recalling the general module form for signal models with shift operators, we have defined $\mathcal{M}_x = \mathbb{C}[x]/p(x)$ to be a regular module with the basis $b_x = (p_0(x), \dots, p_{n-1}(x))$.
+If $x$ is the 1-dimensional shift operator for this signal model, we can naturally extend this to an $n$-dimensional analog with other shift operators $x_0, x_1, \dots, x_{n-1}$ in the module
+\begin{equation}
+    \mathcal{M}_{x_0, \dots, x_{n-1}} = \mathbb{C}[x_0, \dots, x_{n-1}]/\langle p(x_0), \dots, p(x_{n-1})\rangle
+\end{equation}
+Which has a basis that is the cartesian product of all bases of all the 1-dimensional submodules: $b_{x_0, \dots, x_{n-1}} = b_{x_0}\times \cdots \times x_{x_{n-1}}$.
+We shift our focus back to the 2-dimensional signal models pertaining to our processing of images.
+Suppose $x$ and $y$ are our spatial operators over the two axes and each has a dimension corresponding to the number of pixels in that axis, $k$ and $l$, respectively.
+Then we can extend the 1-dimensional signal model on the basis $(p_0, \dots, p_{n-1})$ to
+\begin{equation*}
+    \Phi(\mathbf{s}) = \sum_{0\leq k,l <n}s_{k,l}p_kp_l
+\end{equation*}
+We now take a detour into the formal definition of a tensor product of the two modules $\mathcal{M}_x$ and $\mathcal{M}_y$ and the associated algebra $\mathcal{A}_{x,y}$ thus has the same property:
+\begin{align*}
+    \mathcal{M}_{x,y} &= \mathcal{M}_x\otimes\mathcal{M}_y\\
+    \mathcal{A}_{x,y} &= \mathcal{A}_x\otimes\mathcal{A}_y
+\end{align*}
+This observation gives a valuable insight for the construction of a fast fourier transform in this signal model. Because we are working with a signal whose
+\begin{align*}
+    \Delta : \mathbb{C}[x]&/p(x)\times\mathbb{C}[y]/p(y)\to\\
+    &\bigoplus_{0\leq k,l<n}\mathbb{C}[x]/(x - \alpha_k)\times \mathbb{C}[y](y - \alpha_{l})
+\end{align*}
+This We already know that $\mathcal{M}_x$ and $\mathcal{M}_y$ are modules representing
 
-\subsection{Demonstration}
+The design of a fast 1D DCT goes beyond my understanding, but P\"uschel and other authors have published papers outlining algebraic approaches to its derivation\cite{puschel2003cooley,fastdct1988}.
+Instead, we will use one of many existing libraries with fast DCT-type 2 and DCT-type 3 algorithms.
+The trouble is that most operate exclusively on floating-point numbers which means that the available transforms are not written for our vector space in $(\mathbb{Z}/256\mathbb{Z})^{n^2}$.
+We address this by reinterpreting our vector space in $\mathbb{Q}^n$, performing all of our calculations as rational floating-point numbers, and then casting back down to the original vector space.
+Evidently, this is a cause for a large computation overhead and will almost certainly make our compression system slower than current state-of-the-art systems.
+The importance of matching a signal model to the data being processed is thus emphasized again, though I did not have time to address this issue in my project.
+
+\subsection{Algorithm Design}
+%TODO: explain the layout of the matrix in terms of frequency
+Having designed a two-dimensional DCT that operates in $O(n^2\log{n})$ on floating-point $n\times n$ matrices, we shift our attention to designing a signal processing pipeline for image compression.
+There are two fundamental ideas behind most DCT-based image compression techniques: (1) at most resolutions, lower-frequency components occupy a bulk of the signal energy and perhaps consequentially, (2) higher-frequency components are less important to the characteristics of the image.
+Compression algorithms will find a means of stripping these high-frequency components out of the representation of their signal.
+Some encodings like JPEG do this in a nearly lossless manner by applying a normalization to these frequency components called quantization\cite{JPEG1992}.
+Quantization reduces all but the most powerful high-frequency components to zero leaving a frequency matrix that is very sparse in the lower-right region.
+The JPEG encoding employs a technique called ``zigzag scanning'' to encode these redundancies in a manner that will likely encode most of the zeros contiguously as shown in \ref{fig:zigzag}.
+\begin{figure}[h]
+    \begin{center}
+        \includegraphics[width=2in]{figures/jpeg_zigzag.png}
+    \end{center}
+    \caption{The zigzag scanning used in JPEG intuitively maximizes the number of contiguous encountered in the frequency matrix}
+    \label{fig:zigzag}
+\end{figure}
+In this compression scheme, we take a much simpler approach by simply taking the output matrix of the DCT type-2 and discarding the indices $k,l > m$ for $m \geq 1$ thereby constructing a low-rank approximation of the frequency matrix.
+This compression scheme is bound to have a higher data-loss rate as it simply ignores information that might be outside of this frame of the matrix.
+Relying on the assumption that most of the signal energy is concentrated in the lower spatial frequencies of the signal, this method should be a much better approximation than an arbitrary selection of $m\times m$ matrix elements.
+
+One might be tempted to take this scheme and apply the transformation and rank reduction to the entire image in one two-dimensional DCT type-2.
+This method has two major flaws, the first being that our designed DCT only operates on square matrices while images can be any rectangular shape.
+We would have to either crop every image to a square or derive some type of padding scheme to extend the image bounds to a square.
+The second is that an unnecessary asymptotic complexity degree is used as our algorithm is $\mathcal{O}(n^2\log{n})$.
+
+Instead, we can make our DCT a constant size and apply it to sub-matrices of the image.
+Our asymptotic runtime complexity is now $\mathcal{O}(n^2)$ and we have also somewhat solved the square image problem by reducing the proportions of the image that need to be padded or cropped.
+One question
+
+The described process can be summarized in the following
+
+\tikzstyle{startstop} = [rectangle, rounded corners, minimum width=2cm, minimum height=0.7cm,text centered, draw=black, fill=red!30]
+\tikzstyle{filter} = [rectangle, minimum width=2cm, minimum height=0.7cm,text centered, draw=black, fill=green!50]
+\tikzstyle{arrow} = [thick,->,>=stealth]
+
+\begin{center}
+    \begin{tikzpicture}[node distance=1cm]
+        \node (image) [startstop, xshift=-2cm] {Image};
+        \node (disassemble) [filter, right of=image, xshift=2cm] {Disassemble};
+        \node (dct) [filter, below of=disassemble] {DCT II 2D};
+        \node (reduce) [filter, left of=dct, xshift=-2cm] {Rank Reduce};
+        \node (write) [startstop, below of=reduce, xshift=1cm] {Write Encoding};
+
+        \node (read) [startstop, below of=write, yshift=-0.5cm] {Read Encoding};
+        \node (augment) [filter, below of=read, xshift=-2cm] {Augment Matrices};
+        \node (idct) [filter, right of=augment, xshift=2.5cm] {DCT III 2D};
+        \node (reassemble) [filter, below of=idct] {Reassemble};
+        \node (im) [startstop, left of=reassemble, xshift=-1.5cm] {Image};
+
+        \draw [arrow] (image) -- (disassemble);
+        \draw [arrow] (disassemble) -- (dct);
+        \draw [arrow] (dct) -- (reduce);
+        \draw [arrow] (reduce) -- (write);
+        \draw [arrow] (read) -- (augment);
+        \draw [arrow] (augment) -- (idct);
+        \draw [arrow] (idct) -- (reassemble);
+        \draw [arrow] (reassemble) -- (im);
+    \end{tikzpicture}
+\end{center}
+I implemented this using fftw3 floating point dct transformations and a command-line program written in C.
+It achieves most of the functionality that I was hoping for with a few issues with rectangular dimensioned images.
 We first take an image and convert it into its uncompressed form, in this example, I have taken the following image and converted it into binary ppm:
-\begin{figure}
+\begin{figure}[h]
     \centering

### 2022-05-10 paper.tex
application section

@@
     In this paper, we explore constructs in image processing through the work of Markus P\"uschel on algebraic signal processing theory.
-His main objective was to introduce well-established concepts in signal processing in terms of a comprehensive algebraic theory.
+His main objective was to introduce well-established concepts in signal processing as a comprehensive algebraic theory.
 Here, we do much of the same, in fact a significant portion of this paper is devoted to reviewing essential ideas in his paper.
@@
 For example, an engineer who has significant experience in one field of signal processing may find a natural means of recognizing parallels in an entirely different field.
-It also becomes clear that several tasks in processing research that are seemingly the product of ingenious creativity are actually quite procedural.
-The design of fast transform algorithms, for example, can be as simple as observing the symmetry structure of an underlying signal module.
+It also becomes clear that signal processing optimizations that are seemingly the product of ingenious creativity are actually quite procedural in their construction.
+The design of fast fourier transform algorithms, for example, can be as simple as observing the symmetry structure of an underlying signal module.
 
-The structure of this paper will first explore some of the foundational concepts introduced in the P\"uschel paper, which is primarily an expository section.
+The structure of this paper will first explore the foundational concepts of algebraic signal processing, which is primarily an expository section.
 In this section, we also consider some ideas that were not addressed in this work and potential extensions for P\"uschel's signal processing theory.
 We then move on to a practical application of constructing a signal model for images and a fast transform to be used in an image compression program.
-\section{From First Principles}
+\section{Designing the Signal Model}
 In standard signal processing theory, signals live in a vector space like $\mathbb{C}^n$ over the base field, which gives the direct operations of signal addition and ``scalar multiplication''.
@@
 Compressing data usually comes at the expense of additional computation, which is why the development of efficient compression algorithms is a central focus of this field.
-We focus specifically on lossy compression, in which information is lost through the compression processes (ie.
-the original data cannot be reconstructed from the compressed data) but the majority of useful information is preserved.
+We focus specifically on lossy compression, in which information the original data cannot be reconstructed from the compressed data but the majority of useful information is preserved through the compression process.
 One common lossy compression technique involves decomposing some signal into its irreducible representations and then discarding the less-meaningful representations.
 In this section we look at defining a signal model for digital image data and then develop a lossy compression technique based on an efficient transform derived from this signal model.
-There is also a source code accompaniment for this section which can be found \href{https://github.com/henry-2025/math189-spectral-compression}{here}.
-
 \subsection{Choosing an Encoding}
@@
 Our goal here is to convert between two formats, one compressed and the other uncompressed.
-The uncompressed format is closest to the representation of raw sensor signal data while the compressed format is like one that we might wish to store on the hard drive of a computer.
-One family of encodings that stands out is the Netbpbm format, which encodes an image in a byte table corresponding directly to the pixel indices in the displayed image.
-For example, using the using red green and blue channel ordering for three-tuples, the following pixel array and color block are equivalent representations
+The uncompressed format is closest to the representation of how we might want to interpret the image visually while the compressed format does not have to visually interpretable.
+We can readily apply some change of basis or additional encoding that satisfy its primary objective of being small.
+One family of uncompressed encodings that stands out is the Netpbm format, which encodes an image in a byte table corresponding directly to the pixel indices in the displayed image.
+In this example, we use the RGB channel ordering for three-tuples of bytes.
+The following pixel array and color block are equivalent representations
 
@@
 \end{equation*}
-However, we soon realize see that this model would be flawed because we cannot reach all points in a 2d array from a single spatial operator.
+However, we soon realize see that this model would be flawed because it is not a generating element of the group.
+That is, given a single point in our 2D spatial signal module, we cannot reach all other points by applying this one shift operator.
 Instead, we define two spatial shifts, $q_x$ and $q_y$ and assign their propagations to each of the two dimensions such that
@@
 Recalling the general module form for signal models with shift operators, we have defined $\mathcal{M}_x = \mathbb{C}[x]/p(x)$ to be a regular module with the basis $b_x = (p_0(x), \dots, p_{n-1}(x))$.
-If $x$ is the 1-dimensional shift operator for this signal model, we can naturally extend this to an $n$-dimensional analog with other shift operators $x_0, x_1, \dots, x_{n-1}$ in the module
-\begin{equation}
-    \mathcal{M}_{x_0, \dots, x_{n-1}} = \mathbb{C}[x_0, \dots, x_{n-1}]/\langle p(x_0), \dots, p(x_{n-1})\rangle
-\end{equation}
-Which has a basis that is the cartesian product of all bases of all the 1-dimensional submodules: $b_{x_0, \dots, x_{n-1}} = b_{x_0}\times \cdots \times x_{x_{n-1}}$.
+If $x$ is the 1-dimensional shift operator for this signal model, we can naturally extend this to an $n$-dimensional analog by introducing new shift operators $x_0, x_1, \dots, x_{n-1}$ in the module as we have explained above.
+Assuming we are using the regular module, this means that the module and algebra defined over the polynomials on $x_0, x_1, \dots, x_{n-1}$:
+\begin{align}
+    \mathcal{A}_{x_0,\dots, x_{n-1}} &= \mathcal{M}_{x_0, \dots, x_{n-1}} = \nonumber\\
+    &\mathbb{C}[x_0, \dots, x_{n-1}]/\langle p(x_0), \dots, p(x_{n-1})\rangle\label{eq:module}
+\end{align}
+Which has a basis that is the cartesian product of all bases of all the 1-dimensional submodules: $b_{x_0, \dots, x_{n-1}} = b_{x_0}\times \cdots \times b_{x_{n-1}}$.
 We shift our focus back to the 2-dimensional signal models pertaining to our processing of images.
-Suppose $x$ and $y$ are our spatial operators over the two axes and each has a dimension corresponding to the number of pixels in that axis, $k$ and $l$, respectively.
-Then we can extend the 1-dimensional signal model on the basis $(p_0, \dots, p_{n-1})$ to
+Suppose $x$ and $y$ are our spatial operators over the two axes, then our module definition follows that of \eqref{eq:module} with polynomials over $x$ and $y$.
+
+We now have the facility to define the bijection of our signal model.
+Assume that the maximum degree of each $x$ and $y$ is $n$ (corresponding to $V$ being composed of $n\times n$ matrices).
+We define $p_0, \dots, p_{n-1}$ to be the polynomials of the delta functions for each of the 1D submodules of $M_{x,y}$.
+Then we can extend the 1D signal model on the basis $(p_0, \dots, p_{n-1})$ to
 \begin{equation*}
@@
 \end{equation*}
-We now take a detour into the formal definition of a tensor product of the two modules $\mathcal{M}_x$ and $\mathcal{M}_y$ and the associated algebra $\mathcal{A}_{x,y}$ thus has the same property:
+This completes the basic definition of the signal model.
+We observe that using the usual definitions of the 1D signal models for $\mathcal{M}_x$ and $\mathcal{M}_y$, there exists a bijection between $\mathcal{M}_{x,y}$ and $\mathcal{M}_x\times \mathcal{M}_y$, which means that $\mathcal{M}_{x,y}$ is a \textit{tensor product} of $\mathcal{M}_x$ and $\mathcal{M}_y$.
+Given that we have defined $\mathcal{M}_{x,y}$ as a regular module, this statement also holds for $\mathcal{A}_{x,y}$:
 \begin{align*}
@@
 \end{align*}
-This observation gives a valuable insight for the construction of a fast fourier transform in this signal model. Because we are working with a signal whose
+This observation gives a valuable insight for the construction of a fast Fourier transform in this signal model.
+The decomposition of a $\mathcal{M}_x\times \mathcal{M}_y$ into its irreducible submodules can be written as the following
 \begin{align*}
@@
 \end{align*}
-This We already know that $\mathcal{M}_x$ and $\mathcal{M}_y$ are modules representing
+Which means that Fourier transform on the 2D signal space can be decomposed into 1D Fourier transforms along the axes.
+If $\mathcal{F}$ is such a 1D Fourier transform, then we define $\mathcal{F}_{2D} = \mathcal{F}\times \mathcal{F}$, which we can interpret as an operator that executes the following steps:
+\begin{enumerate}
+    \item Apply $\mathcal{F}$ to each $(s_{r,0}, s_{r,1}, \dots, s_{r,n-1})$ for $0\leq r < n$.
+That is, apply $\mathcal{F}$ to each row of $\mathbf{s}$.
+We refer to the output of this as the matrix $[f_{k,l}]$
+    \item Apply $\mathcal{F}$ te each $(f_{0,r}, f_{1,r}, \dots, f_{n-1,r})$ for $0\leq r < n$.
+That is, apply $\mathcal{F}$ to each column of $[f_{k,l}]$.
+The output of this stage will be the matrix $[g_{k,l}]$.
+This is the output of our 2D FFT
+\end{enumerate}
+We choose $\mathcal{F}$ to be DCT-type 2 for ability to convert purely real signals to a real basis where the submodules are ordered by low-to-high frequency.
+We will see why this choice is convenient in the later portions of our compression pipeline.
+Having performed $2n$ instances of DCT-type 2 and assuming DCT-type 2 is an $\mathcal{O}(n\log{n})$ operation, we perform the 2D fourier transform in $\mathcal{O}(n^2\log{n})$ operations.
+Deriving an inverse operation, we simply apply $\mathcal{F}^{-1}$ in the reverse order as defined for $\mathcal{F}\times\mathcal{F}$.
+DCT-type 2's inverse is defined as DCT-type 3, meaning these are the two essential subroutines that we will use in our algorithm.
 
@@
 The importance of matching a signal model to the data being processed is thus emphasized again, though I did not have time to address this issue in my project.
-
 \subsection{Algorithm Design}
-%TODO: explain the layout of the matrix in terms of frequency
 Having designed a two-dimensional DCT that operates in $O(n^2\log{n})$ on floating-point $n\times n$ matrices, we shift our attention to designing a signal processing pipeline for image compression.
@@
 Compression algorithms will find a means of stripping these high-frequency components out of the representation of their signal.
-Some encodings like JPEG do this in a nearly lossless manner by applying a normalization to these frequency components called quantization\cite{JPEG1992}.
-Quantization reduces all but the most powerful high-frequency components to zero leaving a frequency matrix that is very sparse in the lower-right region.
-The JPEG encoding employs a technique called ``zigzag scanning'' to encode these redundancies in a manner that will likely encode most of the zeros contiguously as shown in \ref{fig:zigzag}.
+Some encodings like JPEG do this in a nearly lossless manner\cite{JPEG1992}.
+First, it applies a normalization to these frequency components called quantization, reducing all but the most powerful high-frequency components to zero.
+The output of the quantization stage leaves a matrix that is very sparse in the lower-right region.
+The JPEG encoding employs a technique called ``zigzag scanning'' to encode these redundancies in a manner that tries to maximize contiguous redundancies as shown in \ref{fig:zigzag}.
 \begin{figure}[h]
@@
     \end{center}
-    \caption{The zigzag scanning used in JPEG intuitively maximizes the number of contiguous encountered in the frequency matrix}
+    \caption{The zigzag scanning used in JPEG intuitively maximizes the number of contiguous zeros encountered in the frequency matrix}
     \label{fig:zigzag}
 \end{figure}
-In this compression scheme, we take a much simpler approach by simply taking the output matrix of the DCT type-2 and discarding the indices $k,l > m$ for $m \geq 1$ thereby constructing a low-rank approximation of the frequency matrix.
+In this compression scheme, we take a much simpler approach by simply taking the output matrix of the 2D DCT type-2 and discarding the indices $k,l > m$ for $m \geq 1$ thereby constructing a low-rank approximation of the frequency matrix.
 This compression scheme is bound to have a higher data-loss rate as it simply ignores information that might be outside of this frame of the matrix.
@@
 
-Instead, we can make our DCT a constant size and apply it to sub-matrices of the image.
+Instead, we can make our DCT a constant size and apply it to a disassembly of sub-matrices of the original signal.
 Our asymptotic runtime complexity is now $\mathcal{O}(n^2)$ and we have also somewhat solved the square image problem by reducing the proportions of the image that need to be padded or cropped.
-One question
+One question still remains regarding the size of these disassembled submatrices.
+We choose $8\times 8$ with no particular mathematical grounding, only noting that JPEG also chooses this size based on hardware constraints and empirical testing \cite{JPEG1992}.
 
-The described process can be summarized in the following
-
+The described process can be summarized in the following flow graphs with decompression being the regular construction of inverse operations:
 \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=2cm, minimum height=0.7cm,text centered, draw=black, fill=red!30]
@@
 
-        \node (read) [startstop, below of=write, yshift=-0.5cm] {Read Encoding};
+        \node[draw] (compression) [above of=image, xshift=1cm] {Compression Pipeline};
+        \node[draw] (decompression) [below of=compression, yshift=-4cm]{Decompression Pipeline};
+
+        \node (read) [startstop, below of=decompression] {Read Encoding};
         \node (augment) [filter, below of=read, xshift=-2cm] {Augment Matrices};
@@
 \end{center}
-I implemented this using fftw3 floating point dct transformations and a command-line program written in C.
-It achieves most of the functionality that I was hoping for with a few issues with rectangular dimensioned images.
-We first take an image and convert it into its uncompressed form, in this example, I have taken the following image and converted it into binary ppm:
+
+\subsection{Implementation and Results}
+I implemented this using FFTW3 floating-point DCT transformations and a command-line program written in C.
+The source code can be found \href{https://github.com/henry-2025/math189-spectral-compression}{here}.
+The encoding, named JHPeg, achieves most of the functionality that I was hoping for with a few issues with rectangular dimensioned images.
+As I anticipated, the use of floating-point math and type conversions makes this program run about 25 times slower than current JPEG encoders.
+However, the resulting files are comparable in size with a compression rate that retains somewhat acceptable image quality as shown in the following table:
+\begin{center}
+    \begin{tabular}{c|c c c}
+        & JPEG & JHPeg & Uncompressed\\
+        \hline
+        Encoding Time & 0.12s & 3.10s & N/A\\
+        File Size & 281K & 338K & 3.6M
+    \end{tabular}
+\end{center}
+It is also helpful to observe the visual effects of low-rank approximation.
+Since the compression scheme is more unrefined than JPEG's we expect to see more inaccuracy in our image outputs.
+In \ref{fig:compression_outputs}, we observe the ``artifacts'' that are produced in the low-rank approximation images versus a state-of-the-art encoding..
+
 \begin{figure}[h]
     \centering
-    \begin{subfigure}[m]{0.3\textwidth}
+    \begin{subfigure}[m]{0.5\textwidth}
         \centering
-        \includegraphics[width=\textwidth]{figures/lion_compressed.jpg}
-        \caption{A high-resolution image with 75\% compression applied}
+        \includegraphics[width=\textwidth]{figures/lion_compare.jpg}
+        \caption{A $960\times 960$ image with 75\% compression applied on the left(we have halved the ranks of the DCT matrices).}
         \label{fig:lion}
@@
     \hfill
-    \begin{subfigure}[b]{0.3\textwidth}
+    \begin{subfigure}[b]{0.5\textwidth}
         \centering
-        \includegraphics[width=\textwidth]{figures/b_compressed.jpg}
-        \caption{A lower-resolution image with ``sharp'' features. Artifacts in the 75\% compression are more evident.}
+        \includegraphics[width=\textwidth]{figures/b_compare.jpg}
+        \caption{A lower-resolution image with ``sharp'' features. Artifacts in the 75\% compression are more evident as seen in the compressed image on the left.}
         \label{fig:letter_b}
     \end{subfigure}
+    \caption{Two different outputs of the compression algorithm that demonstrate its performance.}
     \label{fig:compression_outputs}
-    \caption{Two different outputs of the compression algorithm that demonstrate its performance.}
 \end{figure}
+\subsection{Conclusion}
+Developing the compression scheme was enlightening because it tied together many of the insights from P\"uschel's work in a practical application that I already have a lot of appreciation for.
+He states that the task of deriving fast fourier transform algorithm from basic principles is greatly simplified through an algebraic approach.
+While I do believe this statement, designing an fft from the ground up was still too challenging a task to tackle in a single project.
+I think P\"uschel means that these understandings lead to systematic constructions of fast algorithms rather than their discovery by chance.
+This was certainly evident when I was able to reinterpret existing algorithms and apply them to new tasks.
 \bibliographystyle{plain}

### 2022-05-11 paper.tex
close to done

@@
 
-In the broadest sense, signal processing deals with functions defined on sets, often times countable sets in the context of digital computing.
+Signal processing deals with functions defined on sets, often times countable sets in the context of digital computing.
 We also have a set of functions, commonly referred to as ``filters'' that transform these signals.
@@
 
-To one with experience in one of the fields of signal processing, reading this last sentence should have been fairly straightforward, drawing on some assumptions.
+Some with experience in one of the fields of signal processing may find this last sentence fairly straightforward, drawing on some assumptions.
 If one had contextualized this as the processing of complex digital sensor data, they might have assumed that $x$ and $y$ live in the vector space $\mathbb{C}^n$ over the field $\mathbb{C}$ and that the expression $r_1x + r_2y$ is simply the vector sum of $x$ and $y$ scaled by $r_1$ and $r_2$.
@@
 
-One might question the practical value of an algebraic exploration of already well-established science.
+One might question the practical value of an algebraic exploration of an already well-established science.
 In this inquiry, we find many justifications that apply far beyond the mathematics alone.
 For example, an engineer who has significant experience in one field of signal processing may find a natural means of recognizing parallels in an entirely different field.
-It also becomes clear that signal processing optimizations that are seemingly the product of ingenious creativity are actually quite procedural in their construction.
+It also becomes clear that signal processing optimizations which are seemingly the product of ingenious creativity are actually quite procedural in their construction.
 The design of fast fourier transform algorithms, for example, can be as simple as observing the symmetry structure of an underlying signal module.
 
-The structure of this paper will first explore the foundational concepts of algebraic signal processing, which is primarily an expository section.
-In this section, we also consider some ideas that were not addressed in this work and potential extensions for P\"uschel's signal processing theory.
+This paper will first explore the foundational concepts of algebraic signal processing, which is primarily an expository section.
+We also consider some ideas that were not addressed in this work and potential extensions for P\"uschel's signal processing theory.
 We then move on to a practical application of constructing a signal model for images and a fast transform to be used in an image compression program.
 \section{Designing the Signal Model}
-In standard signal processing theory, signals live in a vector space like $\mathbb{C}^n$ over the base field, which gives the direct operations of signal addition and ``scalar multiplication''.
-This vector space structure also endows us with several other notions like dimension, basis, linear mappings and subspaces.
+\subsection{Definition}
+In standard signal processing theory, signals live in a vector space such as $\mathbb{C}^n$ over a base field, which gives the direct operations of signal addition and ``scalar multiplication''.
+This vector space structure also endows us with several other notions like dimension, basis, linear mappings and subspaces which all interpreted in their unique signal processing contexts.
 We then encounter objects called filters, linear operators that act on signals to produce new signals.
 The operation of a filter on a signal is a different operation than the sum of two signals.
-We use the multiplication operation to represent this action of a filter on a signal
+We use $\cdot$ to represent this action of a filter on a signal.
+Note that this is not a commutative operator as there is no overarching practice of applying signals to filters.
 \begin{equation*}
@@
 \end{equation*}
-filtering also obeys the laws of distributivity
+Filtering also obeys the laws of distributivity
 \begin{align*}
@@
 \end{align*}
+Let us define the sets of all filters $\mathcal{A}$ and all signals $\mathcal{M}$ in an arbitrary signal processing context.
 For $h,h'\in \mathcal{A}$ and $s,s'\in\mathcal{M}$ and $\alpha$ in the base field, we also have
@@
 \end{align*}
-and filters can be operated on in various manners as well
+Filters can also operate one another and can be scaled in the base field. A parallel filter application and a series filter application are the two different configurations in which this can happen:
 \begin{align*}
@@
 \end{align*}
-Beyond these properties, it must be mentioned that scalar multiplication of filters has left and right-distributivity and that multiplication by scalars is fully commutative with filter multiplication.
-At this point, we see that the structure of this system can be defined using formal mathematical language.
-Our set of filters forms an algebra, which we refer to as $\mathcal{A}$.
-Our vector space is somehow operated on by $\mathcal{A}$, meaning it must have an associated $\mathcal{A}$-module called $\mathcal{M}$.
+It must also be mentioned that scalar multiplication of filters has left and right-distributivity and that multiplication by scalars is fully commutative with filter multiplication.
+
+We now see that $\mathcal{A}$ and $\mathcal{M}$ can be defined using formal mathematical language.
+Our set of filters forms an algebra over the base field.
+The set of signals is an associated $\mathcal{A}$-module.
 However, it must be clearly stated that $\mathcal{M}$ does not necessarily equal $V$ as the operation of $\mathcal{A}$ on $\mathcal{M}$ is not strictly defined as an operation on a vector space.
@@
 This is the central device in signal processing theory: a \textit{signal processing model} on a vector space $V$ consists of an ordered tuple $(\mathcal{A}, \mathcal{M}, \Phi)$ of algebraic objects as defined above.
-Through this model, we can exploit all of the mathematics of representation theory, harmonic analysis, and
+Through this model, we can exploit representation theory, harmonic analysis, and abstract algebra to explore the mathematics of signal processing.
 
 One may note that as soon as $\Phi$ is defined, we have also given definitions for all of $V, \mathcal{M}$ and $\mathcal{A}$, meaning a signal model can be tersely expressed as this bijective function $\Phi$.
+\subsection{Model Derivation}
+In this section, we show that $\mathcal{A}$ and $\mathcal{M}$ can be defined entirely when given just a \textit{shift operator} and a \textit{boundary condition}.
+Signal processing applications typically associate their vector spaces with \textit{samples}, entries in each vector that have an association with some measurable dimension.
+The two most common examples of these dimensions are time and space, where one might imagine sampling sunlight intensity throughout the day or .
+
+We first focus on these time-dimension signal models.
+In such a model, each sample in a vector $\mathbf{s} = [s_0, s_1, \cdots, s_n]^T$ has an associated time index $t_i$, which offers an interpretation of our signal as
+\begin{equation}
+    \sum_{k=0}^ns_kt_k\label{eq:signal1}
+\end{equation}
+The usual extension of this finite sum to a series applies to infinite sample sequences.
+Consider the fact that time has an inherent direction, such that we can only progress forward.
+For a sequence of time indices $t_0, t_1, \dots$, the only reasonable progression through the sequence is from $t_0$ to $t_1$ to $t_2$ and so forth.
+In this manner, we consider a shift operator $q$ to be a means of progressing through time indices by operating on them.
+$q$ applies itself to a time index $t_i$ through $\diamond$ such that
+\begin{equation*}
+    q\diamond t_i = t_{i+1} 
+\end{equation*}
+Surprisingly, this operation alone allows us to define a large portion of our signal model.
+If we let $q = x$ and $t_0 = 1$, then the recurrence $t_{n+1} = xt_n$ is solved by $t_n = x^n$.
+This then reduces \eqref{eq:signal1} to a sum
+\begin{equation*}
+    \sum_{k}s_kx^k
+\end{equation*}
+In general, for any time-sequence model (finite or infinite), the above equation defines our signal model bijection $\Phi(\mathbf{s})$ while placing certain constraints on $V$.
+It does not, however strictly define $\mathbf{M}$ until we can specify the boundary condition.
+For our purposes we consider a finite dimensional $V$ in $\mathbb{C}^n$, such that $\Phi$ produces polynomials of degree at most $n-1$.
+The problem now arises that $\mathbb{C}[x]$ for finite length could not be a module in any algebra as it is not closed under multiplication nor is any other finite length polynomial set.
+For example, $x^{n-1},x\in \mathbb{C}[x]$ but $x^{n-1}x = x^n\not\in \mathbb{C}[x]$.
+Thus arises the need for a \textit{boundary condition}, simply a means of expressing $x^n$ as a polynomial of degree at most $n-1$.
+A boundary condition takes the most general form
+\begin{equation}
+    x^n = r(x) = \sum_{k=0}^{n-1}\beta_kx^k
+\end{equation}
+As a result, polynomials $x^{k+n}$ are equivalent to some $r_k(x)\mod{(x^n-r(x))}$, which tells us that instead of living in $\mathbb{C}[x]$, our polynomials live in $\mathbb{C}[x]/(x^n-r(x))$.
+Indeed, the operation of multiplication $\mod{(x^n-r(x))}$ is closed and thus can form a module.
+
+Our definition of a boundary condition and a shift operator have now allowed us to generally define $\Phi$ and $\mathcal{M}$ in our signal model and we will soon show how they offer suggestions for $\mathcal{A}$.
+However, we must first focus the choice of $r(x)$, as it determines several additional important properties about the signal model:
+\begin{enumerate}
+    \item If $x|r(x)$ then $x$ divides $(x^n - r(x))$ and thus $x$ is not invertible $\mod{(x^n - r(x))}$.
+Thus the operators $x^{-1}, x^{-2}, \dots$ do not exist and we say that the past is inaccessible in our model
+    \item $x\nmid r(x)$ then $x^{-1}$ does exist, in particular $x^{-1} = -\frac{1}{\beta_0}(\beta_1 + \beta_2x + \cdots +\beta_{n-1}x^{n-2} - x^{n-1})$ and this is the machinery used to access the past up to $x^{1-n}$.
+\end{enumerate}
+This ``accessing of the past'' might seem to violate the concept of time shift which processes in only one direction.
+Note that the existence of $x^{-1}$ does not imply that moving from $t_0$ to $t_{-1}$ is possible, it only means that $t_{-1}$ exists and behaves as would any other time index such that $q\diamond t_{-1} = t_0$.
+This notion becomes clearer when we discuss filtering shortly.
+It finally stands to remark at the fact that our choice of $r(x)$ determines what happens at \textit{both} the left and right boundaries of polynomials in $\mathcal{M}$.
+
+We now consider the conventional notion of \textit{shift-invariance} in signal processing and this to determine one particular candidate for $\mathcal{A}$.
+An algebra of filters $\mathcal{A}$ is shift-invariant if for $h\in \mathcal{A}$ and a signal $s$ in the associated $\mathcal{A}$-module $\mathcal{M}$, the shift operator commutes with $h$:
+\begin{equation*}
+    h(xs) = xh(s) \iff hx = xh
+\end{equation*}
+Where $xs$ is the action of applying $x$ to each time index in $s$.
+Noting this property that $\mathcal{A}$ must commute under $x$, a very natural choice is to assign $\mathcal{A} = \mathcal{M}$.
+Indeed, the polynomial algebras are a valid candidate for the filter space of time-invariant filters.
+They have the additional convenient property that they allow for a very easy algebraic interpretation of a finite impulse response (FIR) filter.
+For an example, suppose we have some filter $h = 1 + x\in \mathcal{A} = \mathbb{C}[x]/(x^{9} - 1)$.
+In this case, applying $h$ to a signal $s\in \mathbb{M} = \mathcal{A}$ is equivalent to saying.
+\begin{align*}
+    s' &= hs\\
+    s'_k &= \begin{cases}
+        s_0 - s_9 & k = 0\\
+        2s_9 + s_8 & k = 9\\
+        s_{k-1} + s_k & \text{otherwise}       
+    \end{cases}
+\end{align*}
+which demonstrates the FIR behavior that one might expect, especially in the third case.
+
+We finally come back to this notion of access to the past in terms of filters.
+When $x^{-1}, \dots, x^{1-n}$ exist, then filters can also draw on information that exists in the opposite time direction.
+While we have defined this as the ``past'' in our signal module, note that when $x^{-1}$ exists, filters can be characterized as non-causal.
+That is they have sight into impulses from both time directions.
+Causal filters, on the other hand, only have sight into the past and we thus delineate between these two in the algebraic sense based on $x^{-1}$'s existence.
 
 \subsection{Fourier Transforms}
-Given that we have now associated the signal space with a module, we can define Fourier transforms on our signal space.
+Given that we have now associated the signal space with a module, the construction of a Fourier transforms on the signal space can be carried out using the algebraic definitions encountered in class.
 For an arbitrary $\mathcal{A}$-module $\mathcal{M}$, let $M$ denote the set of all irreducible $\mathcal{A}$-submodules.
-In the general case, we cannot be certain that $\mathcal{M}$ can be decomposed into $M$ but in the case that
-\begin{equation*}
-    \mathcal{M} = \bigoplus_{\omega\in M}\omega
-\end{equation*}
+In the general case, we cannot be certain that $\mathcal{M}$ can be decomposed into $M$ but in the case that it can we know that
+\begin{equation}
+    \mathcal{M} = \bigoplus_{\mathcal{M}_k\in M}\mathcal{M}_k\label{eq:decomp}
+\end{equation}
 we have the machinery to construct a Fourier Transform for our signal model.
-Once we choose bases for the irreducible representations of $\mathcal{M}$ allows us to coordinatize this transformation, thus offering a linear representation which we will call $\mathcal{F}$.
+Once we choose a basis $b_0, b_1, \dots, b_n$ for the irreducible representations of $\mathcal{M}$ we are able to coordinatize this transformation by projecting each of these basis elements into the submodules as shown in \eqref{eq:decomp}.
+The linear representation of this transformation is then offered by the matrix formed by each of these coordinate decompositions $\mathcal{F}$.
 
@@
 The source code can be found \href{https://github.com/henry-2025/math189-spectral-compression}{here}.
-The encoding, named JHPeg, achieves most of the functionality that I was hoping for with a few issues with rectangular dimensioned images.
+The encoding, named JHPeg, achieves most of the functionality that I was hoping for with a few issues with rectangular-dimensioned images.
 As I anticipated, the use of floating-point math and type conversions makes this program run about 25 times slower than current JPEG encoders.
@@
 Since the compression scheme is more unrefined than JPEG's we expect to see more inaccuracy in our image outputs.
-In \ref{fig:compression_outputs}, we observe the ``artifacts'' that are produced in the low-rank approximation images versus a state-of-the-art encoding..
+In \ref{fig:compression_outputs}, we observe these ``artifacts'' that are produced in the low-rank approximation images versus a state-of-the-art encoding..
 
@@
         \includegraphics[width=\textwidth]{figures/b_compare.jpg}
-        \caption{A lower-resolution image with ``sharp'' features. Artifacts in the 75\% compression are more evident as seen in the compressed image on the left.}
+        \caption{A $256\times 256$ image with sharp edge features. Artifacts in the 75\% compression are more evident as seen in the compressed image on the left.}
         \label{fig:letter_b}

### 2022-05-11 paper.tex
final submission

@@
 \usepackage{tikz}
+\usepackage{amsthm}
 
@@
 }
+\newtheorem{theorem}{Theorem}[section]
 
@@
 \end{equation*}
-Surprisingly, this operation alone allows us to define a large portion of our signal model.
+This operation alone allows us to define a large portion of our signal model.
 If we let $q = x$ and $t_0 = 1$, then the recurrence $t_{n+1} = xt_n$ is solved by $t_n = x^n$.
@@
 In general, for any time-sequence model (finite or infinite), the above equation defines our signal model bijection $\Phi(\mathbf{s})$ while placing certain constraints on $V$.
-It does not, however strictly define $\mathbf{M}$ until we can specify the boundary condition.
+It does not, however strictly define $\mathcal{M}$ until we can specify the boundary condition.
 For our purposes we consider a finite dimensional $V$ in $\mathbb{C}^n$, such that $\Phi$ produces polynomials of degree at most $n-1$.
@@
     \item If $x|r(x)$ then $x$ divides $(x^n - r(x))$ and thus $x$ is not invertible $\mod{(x^n - r(x))}$.
-Thus the operators $x^{-1}, x^{-2}, \dots$ do not exist and we say that the past is inaccessible in our model
+The operators $x^{-1}, x^{-2}, \dots$ do not exist and we say that the past is inaccessible in our model
     \item $x\nmid r(x)$ then $x^{-1}$ does exist, in particular $x^{-1} = -\frac{1}{\beta_0}(\beta_1 + \beta_2x + \cdots +\beta_{n-1}x^{n-2} - x^{n-1})$ and this is the machinery used to access the past up to $x^{1-n}$.
@@
 This ``accessing of the past'' might seem to violate the concept of time shift which processes in only one direction.
-Note that the existence of $x^{-1}$ does not imply that moving from $t_0$ to $t_{-1}$ is possible, it only means that $t_{-1}$ exists and behaves as would any other time index such that $q\diamond t_{-1} = t_0$.
+Note that the existence of $x^{-1}$ does not imply the action of moving from $t_0$ to $t_{-1}$, it only means that $t_{-1}$ exists and behaves as would any other time index such that $q\diamond t_{-1} = t_0$.
 This notion becomes clearer when we discuss filtering shortly.
@@
 
-We now consider the conventional notion of \textit{shift-invariance} in signal processing and this to determine one particular candidate for $\mathcal{A}$.
+We now consider the conventional notion of \textit{shift-invariance} in signal processing and use this to determine one particular candidate for $\mathcal{A}$.
 An algebra of filters $\mathcal{A}$ is shift-invariant if for $h\in \mathcal{A}$ and a signal $s$ in the associated $\mathcal{A}$-module $\mathcal{M}$, the shift operator commutes with $h$:
@@
 They have the additional convenient property that they allow for a very easy algebraic interpretation of a finite impulse response (FIR) filter.
+Each sample in the output sequence of an FIR filter is a linear combination of samples in the input sequence over sime time frame.
 For an example, suppose we have some filter $h = 1 + x\in \mathcal{A} = \mathbb{C}[x]/(x^{9} - 1)$.
@@
     s'_k &= \begin{cases}
-        s_0 - s_9 & k = 0\\
-        2s_9 + s_8 & k = 9\\
+        s_0 + s_9 & k = 0\\
         s_{k-1} + s_k & \text{otherwise}       
@@
 \end{align*}
-which demonstrates the FIR behavior that one might expect, especially in the third case.
+which demonstrates the FIR behavior that one might expect, especially in the second case.
+We also observe concretely that our defined boundary condition $r(x) = 1$ simply loops this signal back onto itself, which seems like a familiar construction.
+In fact, this boundary condition choice makes $\mathbb{C}[x]/(x^n - r(x))$ isomorphic to the group algebra of cyclic groups $\mathbb{C}[C_n]$ which we have dealt with many times in Harmonic Analysis on Finite Groups.
 
@@
 When $x^{-1}, \dots, x^{1-n}$ exist, then filters can also draw on information that exists in the opposite time direction.
-While we have defined this as the ``past'' in our signal module, note that when $x^{-1}$ exists, filters can be characterized as non-causal.
-That is they have sight into impulses from both time directions.
+While we have defined this as the ``past'' in our signal module, realize that the direction of time is inverted in the filter algebra.
+That is, $x^{-1}$ grants a filter access to future time indices relative to $t_0$ making them \textit{non-causal}.
 Causal filters, on the other hand, only have sight into the past and we thus delineate between these two in the algebraic sense based on $x^{-1}$'s existence.
@@
 Once we choose a basis $b_0, b_1, \dots, b_n$ for the irreducible representations of $\mathcal{M}$ we are able to coordinatize this transformation by projecting each of these basis elements into the submodules as shown in \eqref{eq:decomp}.
-The linear representation of this transformation is then offered by the matrix formed by each of these coordinate decompositions $\mathcal{F}$.
+The linear representation of this transformation is then offered by the matrix $\mathcal{F}$ formed by each of these coordinate decompositions as column vectors.
 
-\subsection{Shifts and the Z-Transform}
+A natural example of this can be demonstrated with a specific instance of the time-shift model that we were working with.
+We select a special boundary condition where $r(x) = 1$ and thus our polynomial algebra can be expressed as the following
+\begin{align}
+    \mathbb{C}[x]&/(x^{n}-1) \nonumber\\
+    &= \mathbb{C}[x]/(x - 1)(x-\omega)\cdots(x-\omega^{n})\label{eq:factoring}
+\end{align}
+where $\omega$ is the primitive $n$th root of unity $e^{2\pi i/n}$.
+As an aside, we consider the Chinese remainder theorem for polynomials which states that if $p(x)\in \mathbb{C}[x]$ and $p(x) = r(x)s(x)$, then
+\begin{equation}
+    \mathbb{C}[x]/p(x)\cong\mathbb{C}[x]/r(x)\oplus\mathbb{C}[x]/s(x)\label{eq:crt}
+\end{equation}
+Applying \eqref{eq:crt} inductively to \eqref{eq:factoring} allows us to say that $\mathbb{C}[x]/(x^n-1)$ decomposes into
+\begin{equation*}
+    \mathbb{C}[x]/(x-1)\oplus\mathbb{C}[x]/(x-\omega)\oplus\cdots \oplus\mathbb{C}[x]/(x-\omega^{n-1})
+\end{equation*}
+And as each of these is a degree 1 polynomial module, they must all be irreducible and thus form a \textit{spectrum} to be used in our Fourier transform.
 
+In the future, we will assume that $(x^n - r(x))$ is \textit{decomposable} and can be decomposed into the product of degree 1 polynomials, all with multiplicity 1.
+This gives the above expression a general form
+\begin{align}
+    \mathbb{C}[x]&/(x^n-r(x)) = \mathbb{C}[x]/(x-\alpha_0)\nonumber\\
+    &\oplus\mathbb{C}[x]/(x-\alpha_1)\oplus\cdots \oplus\mathbb{C}[x]/(x-\alpha^{n-1})\label{eq:factoring}
+\end{align}
+
+From this, the construction of the standard DFT matrix on $n$-sample data in the cyclic group becomes the procedural task of expressing all of the basis elements $(1, x, \dots, x^{n-1})$ in terms of the direct sum spectrum in \eqref{eq:factoring}.
+This gives us the usual DFT matrix $\text{DFT} = [\omega^{ij}]_{0\leq i, j < n}$.
+\subsection{Z-Transform and the Convolution Theorem}
 Anyone who has focused on control systems or electronics in engineering is likely familiar with the concept of transfer functions on signals and filters represented in the z-domain.
@@
 \end{equation*}
-as we note, one of the convenient properties of such a transform is that the multiplication of signals and filters in the z-domain is simple polynomial multiplication.
-This property is often given without any other context, but here we can trivially justify why this is the case.
+Up until this point, we have simply discussed the z-transform as a means of expressing $\mathbf{s}\in V$ as an element of the signal module $\mathcal{M}$.
+Many engineers note that the above transform allows one to express signals that live in $V = \mathbb{C}^n$ in the same closed-form continuous functions as the filter algebra.
+It conveniently frames filter application and as the multiplication of closed-form functions in z-domain.
+The insight that the our choice of module creates properties that have practical use is another reassuring justification for the framing of signal processing through algebraic theory.
 
-\section{Space and Time Signals}
+In many cases, however, there is not significant choice over the exact algebras and modules that are used.
+The module can sometimes make the process of filtering tedious or it may not inherently lead to a convenient method of filter application.
+For example, working with the polynomial algebras of degree $n$ means that one must multiply $\mathcal{O}(n^2)$ elements when applying a filter to a signal.
+Quadratic scaling for such operations is certainly not ideal.
 
-\section{Deriving Transforms from a Signal Model}
+We then call upon representation theory to attempt to express filters in terms of bases that make multiplication simpler.
+We first assume that our module can be decomposed into a series of irreducible submodules as in \eqref{eq:decomp} and then perform such a decomposition on a filter or signal according to \eqref{eq:factoring}
+\begin{equation*}
+    \mathbf{s} = \sum_{k=0}^{n-1}\alpha_k(s_k)
+\end{equation*}
+Given that each of these terms is a projection into an irreducible submodule of $\mathbb{C}[x]/(x^n-r(x))$, representation theory gives us the insight that these can a all be represented as diagonal matrices in $\mathbb{C}^1$.
+The constructed representation of $\mathbf{s}$ in this basis is then a diagonal matrix in $\mathbb{C}^n$, which establishes the \textit{convolution theorem}.
+\begin{theorem}[Convolution]
+    \label{thm:convolution}
+    Let $s\in \mathcal{M}$ and $h\in \mathcal{A}$ and be a signal and filter $\text{mod} (x^n-a)$.
+Let $\phi$ be a representation of $\mathcal{A}$.
+An arbitrary fourier transform $\mathcal{F}$ has the property
+    \begin{align*}
+        h\cdot s\mod{(x^n-r(x))} \equiv \phi(h) \cdot \mathbf{s}\\
+        \equiv (D\mathcal{F})^{-1}((\mathcal{F}\mathbf{h}) \odot (\mathcal{F}\mathbf{s}))
+    \end{align*} 
+    Where $\mathbf{h}$ and $\mathbf{s}$ are the coefficient vectors of $h$ and $s$ with respect to the common basis $b$ of $\mathcal{A}$ an $\mathcal{M}$.
+$\odot$ is pointwise multiplication of the diagonals of two matrices.
+The matrix $D$ is diagonal and is determined by our chosen basis $b$.
+The P\"uschel paper goes into better detail on why this diagonal matrix is necessary.
+\end{theorem}
+The obvious takeaway from \ref{thm:convolution} is that when the direct sum decomposition properties of $\mathcal{M}$ and $\mathcal{A}$ are satisfied, filter application takes $\mathcal{O}(n)$ operations, a drastic speedup over the arbitrary case.
+\section{A Brief Exploration of Space Signals}
+The signal models that we have realized so far have been time models, but many applications in signal processing cannot be modeled by this oriented time axis.
+Space, for example, has no intrinsic direction and is certainly the most common deviation from the time model in signal processing.
+We consider the vector space $V = \mathbb{C}^\mathbb{Z}$ and define similar space indices $t_n$ and a space shift operator $q$.
+To develop an intuition for how the space operator should behave when applied to $t_n$ we consider a propagation outward from a point in space.
+Where time propagates in one direction, a signal propagating through space from a point should travel in all directions.
+The space model then looks like the following
+\begin{equation*}
+    q\diamond t_n = (t_{n+1} + t_{n-1})/2
+\end{equation*}
+for $n\in \mathbb{Z}$. It is perhaps more convenient to visualize how this relates to the intuition in \ref{fig:space}.
+\begin{figure}[h]
+    \begin{center}
+        \includegraphics[width=2.5in]{figures/space_shift.png}
+        \caption{The space shift operator for a single shift. Source: \cite{AlgebraicSignalProcessing2006}}
+        \label{fig:space}
+    \end{center}
+\end{figure}
+A natural extension from $q$ to the $k$-fold shift operators $q_k$ is the following
+\begin{equation*}
+    q_k\diamond t_n = (t_{n+k} + t_{n-k})/2
+\end{equation*}
+Here, we can observe two important differences between the space and time signal models.
+Most notably, $q_k \neq q^k$ as one can verify by applying $q^2$ to $t_n$.
+Additionally, $q_k = q_{-k}$ which reflects the fact that there is no directionality in the space shift.
+We find that instead, the shift operators are the Chebyshev polynomials of the first kind defined by the recurrence $T_k(q) = 2qT_k(q) - T_{k-1}(q)$
 
-\cite{AlgebraicSignalProcessing2006}
-Basic constructs in signal processing theory are filters, the z-transform, signals and the fourier transform.
-When we view them as they are classically taught, we see filters as linear operators and signals as a vector space, but this paper employs an algebraic perspective to show that these are more than
+\begin{theorem}[Space Shift Operator]
+    The $k$-fold space shift operator is given by $T_k(q)$
+    \begin{proof}
+        By induction on $k$, let $q_0 = 1$ and $q_1 = q$. We have
+        \begin{align*}
+            q_{k+1}\diamond t_n &= (t_{n+k+1} + t_{n-k-1})/2\\
+            &= (t_{n+k+1} + t_{n+k-1} + t_{n-k+1} + t_{n-k-1})/2\\
+            &\quad- (t_{n+k-1} + t_{n-k+1})/2\\
+            &= 2q\diamond(t_{n+k} + t_{n-k})/2\\
+            &\quad- (t_{n+k-1} + t_{n-k+1})/2\\
+            &= (2qq_k - q_{k-1})\diamond t_n
+        \end{align*}
+        which is indeed the recurrence relation associated with Chebyshev Polynomials of the first kind.
+    \end{proof}
+\end{theorem}
+Deriving the module and algebra, the boundary conditions, and the Fourier transform associated with space models follows in almost identical fashion to the construction in time models.
+The process is slightly more involved and results in a much larger space of transforms and boundary conditions than were observed with time signals.
+In fact, we find that there are 16 types of finite bijections $\Phi$ that arise from our choice of boundary conditions.
+There are then 16 associated Fourier transforms known as the discrete cosine and sine transforms that apply to each of these bijections\cite{AlgebraicSignalProcessing2006}.
 
-Z-transform converts a discrete time signal  into a complex frequency-domain representation, which can be thought of as the discrete time analogue of the Laplace transform. The
-
-Define this notion of ``linear time invariance''
-
-However, simple observation shows that this space is not closed under multiplication, the operation corresponding with successive filter application.
-As such, many engineering conventions treat this as multiplication of polynomials modulo $z^{-n}-1$
-
-\subsection{Signal Shift Operator}
-Every signal model that has the shift invariance property has a commutative filter algebra $\mathcal{A}$.
-\section{Further Considerations: Transformations between Algebras}
-Many times in engineering we encounter examples where a transformation may not transform to the same algebra and may not be invertible.
-For example, the process of downsampling a signal (commonly referred to as applying a decimating filter) applies a transformation that reduces the number of samples in a vector.
-This can be thought of as applying a filter with a large null space to a signal.
-
+In-depth derivation of each of these is beyond the scope of this paper, but a central conclusion is that the discrete trigonometric transforms are the applicable Fourier transforms to spatial signals.
+We will apply this insight in the following section.
 \section{Practical Example: Spectral Image Compression}
@@
 Our asymptotic runtime complexity is now $\mathcal{O}(n^2)$ and we have also somewhat solved the square image problem by reducing the proportions of the image that need to be padded or cropped.
-One question still remains regarding the size of these disassembled submatrices.
+One question still remains regarding the size of these disassembled sub-matrices.
 We choose $8\times 8$ with no particular mathematical grounding, only noting that JPEG also chooses this size based on hardware constraints and empirical testing \cite{JPEG1992}.
